\plaintitle

### 2020-03-27 wip-design/paper.tex
Related work

\numberofauthors{3}
\author{%
  \alignauthor{Leave Authors Anonymous\\
    \affaddr{for Submission}\\
    \affaddr{City, Country}\\
    \email{e-mail address}}\\
  \alignauthor{Leave Authors Anonymous\\
    \affaddr{for Submission}\\
    \affaddr{City, Country}\\
    \email{e-mail address}}\\
  \alignauthor{Leave Authors Anonymous\\
    \affaddr{for Submission}\\
    \affaddr{City, Country}\\
    \email{e-mail address}}\\
}

@@

Evaluating ui systems research
\cite{evaluating}

Type providers
\cite{inforich,fsdata,dotdriven}

Assistants for SW / auto-complete
\cite{assistants}

Journalism tools
\cite{idyll}

Notebook systems
\cite{jupyter} or more transparency \cite{wrattler}
managing messes \cite{messes}

Visual data exploration tools
\cite{control,tableau,vizdom}

Wrangler
\cite{wrangler}

Victor
\cite{learnable,principle}

Live and exploratory programming
\cite{review} - review of the two
\cite{exploratory}, which led to new tools~\cite{variolite} (supports informal versioning),

\cite{liveroad}
review the use of live programming in a Smalltalk derived environment. Lighttable~\cite{lighttable}

A more principled approach can be used by systems based on structured editing~\cite{structure-based,livenut,lamdu}

Subtext
\cite{subtext,directprog}

sketch and sketch
\cite{sketchnsketch}

lower this barrier to valid, replicablestatistical analysis, we introduce Tea1, a high-level declar-ative language and runtime system
(still JSON, but good aims)
\cite{tea}

DS.js: Turn Any Webpage into an Example-Centric Live Programming Environment for Learning Data Science
(cool but just JS)
\cite{dsjs}

Eviza: A Natural Language Interface for Visual Analysis
(from Tableau; uses auto-complete to suggest natural language phrases)
\cite{eviza}

Declarative Interaction Design for Data Visualization
(interactive construction of vega-like grammar of graphics)
\cite{interactionviz}

programming by demonstration
\cite{watch}

"Dontcheva et al. [4] use PBD techniques to enable
extraction, integration, and templated search of web data.
Gulwani [8] introduces an algorithm for learning expres-
sive string processing programs from input-output examples"

programming by example as an alternative
\cite{spreadsheetpbe,flashextract}

Lyra: An interactive visualization design environment
\cite{lyra}

end-user programming
\cite{enduser} - overview, \cite{smallmatter} book

touch based data science??? must inevitably have fewer intraections
NA

autocomplete / predictive text work
\cite{predictive} - predicting data transformation scripts in a somewhat more complex language
\cite{proactive} -

\cite{potter}
Potter’s Wheel: An Interactive Data Cleaning System
n interactive data clean-ing system that tightly integrates transformation anddiscrepancy detection.   Users  gradually build  trans-formations  to  clean  the  data  by  adding  or  undoingtransforms on a spreadsheet-like interface; the effectof a transform is shown at once on records visible onscreen.

@@

\textbf{Visual tools.}

\textbf{Programming tools.}
Notebooks

Both follow from assistant tools for software development \cite{assistants}. Our implementation,
outlined in the previous section, implements iterative prompting through a code editor mechanism
designed for code completion. Finally, the work on type providers \cite{inforich,fsdata,dotdriven},
which we extend, also utilizes code completion.

\textbf{Journalism.}
Idyll \cite{idyll}

\textbf{Type providers.}
PL work

>>>>

\numberofauthors{1}
\author{%
  \alignauthor{Leave Authors Anonymous\\
    \affaddr{for Submission}\\
    \affaddr{City, Country}\\
    \email{e-mail address}}
}

@@

The Gamma brings together recent programming language innovations for working with data
\cite{inforich}, work on making programming accessible to non-programmers \cite{enduser,smallmatter}
and work on tools for journalists such as Idyll \cite{idyll}. Related work includes a range of
visual and programmatic tools used by data scientists, work that aims to simplify programming
or explore alternative paradigms and various predictive tools that assist users in completing
programming or data exploration tasks.

% Evaluating ui systems research
% \cite{evaluating}

\subsection{Data science tools}
Data exploration is typically done using spreadsheets, comprehensive visual tools or
via code using notebooks. Notebooks such as Jupyter \cite{jupyter} are also popular with technically
skilled journalists as they allow combining code with commentary and visual outputs. However,
notebooks can lead to messy hard to reproduce code \cite{messes,wrattler}.

Visual data analytics tools \cite{control,vizdom} such as Tableau \cite{tableau} do not require
programming skills, but require mastering a range of different interactions. They also make it
hard to see what data transformations have been done. Several systems for data exploration,
e.g.~Potter's Wheel and Wrangler \cite{potter,wrangler}, and visualization, e.g.~Lyra \cite{lyra},
provide a GUI, but record a transparent script that can be seen and modified by the user.

\subsection{Easier programming tools}
We aim to design a programming environment that is easy to use and learn. Many approaches to this
goal have been tried. Victor \cite{learnable,principle} introduced design principles that
inspired many to build live programming systems \cite{review,liveroad,lighttable} that give
immediate feedaback to help programmers understand how code relates to output and
exploratory systems \cite{variolite,exploratory} that assist with completing open-ended tasks.

To avoid difficulties with editing code as text, some systems use structured editors~\cite{structure-based,livenut,lamdu}.
In Subtext \cite{subtext,directprog} the language itself is co-designed with the editor to make
the interactions with code more natural. Finally, programming can be made easier by designing
high-level declarative abstractions. Tea does so for statistical analyses \cite{tea} and
Satyanarayan et al. \cite{interactionviz,vegalite} does so for interactive data visualizations.

The Gamma is live in that it gives an instant preview of the results. It is text-based,
although it is syntactically simple. The code written in The Gamma is high-level and so work
on declarative abstractions is a valuable inspiration.

\subsection{Code completion and assistants}
A key component in The Gamma design is the use of auto-complete for offering possible
operations. Code completion has been pioneerd by Kaiser~\cite{assistants}, but our work is more
directly inspired by the work on type providers \cite{inforich,fsdata,dotdriven}, which utilize
code completion to integrate external data into a statically typed programming language.

When using type providers, completions are offered as the user types code. However, completion
tools can also recmmend scripts when users begin interacting with data \cite{predictive,proactive}.
Alternatively, they can also be guided by natural language as for example Eviza~\cite{eviza},
which uses natural language for data exploration. CodeMend~\cite{codemend} uses natural language
for more general code edits, but uses data visualization as a case study.
Finally, data entry interfaces based on completion such as Dasher \cite{dasher} utilize an
interaction principle similar to our iterative prompting.

\subsection{Programming without writing code}

A number of tools for data exploration use alternative approaches to programming where
the user does not write code. In programming by example \cite{byexample}, the user gives
examples of desired results. In data science, this approach has been used for example for
specifying data transformations in spreadsheets and data extraction \cite{spreadsheetpbe,flashextract}.

In direct manipulation \cite{direct}, a program is specified by directly interacting with the
output. This has been used in the visual domain \cite{sketchnsketch}, but also for data querying
\cite{dynamicq,vlang}. The VQE language~\cite{visage} also considers how to allow code reuse and
modification in this context. Direct manipulation can also support data exploration by letting
users partially edit queries,~e.g. by changing quantifiers as in DataPlay~\cite{dataplay}.

@@
